Realizing Temporal Graphs

### 2023-01-18 RealizingTemporalGraphs.tex
Changes, Tuesday 17th January

Note that we can easily solve the \deltaExact\ by guessing the structure of temporal paths among all pairs of vertices.
But, this would create $f(n,m)$ guesses, and the running time of the algorithm would be too big.
Therefore, we approach the problem using some special properties, which allow us to create $f(k)$ different guesses.
We solve each guess by creating an ILP instance with $f(k)$ variables and $O(n^c)$ constraints, for some $c \in \mathbb{N}$.
The ILP instance is a \textsc{Yes} instance if and only if $D$ is a \textsc{Yes} instance of \deltaExact.
Moreover, from a positive solution of the ILP instance we can construct the desired labeling $\lambda$.
Using results from \cite{Lenstra1983Integer} we know that an ILP instance can be solved in FPT time with respect to the number of variables.
Therefore we can solve our ILP instance in $f(k)$ time.

@@

We use the above results in the following way.
Let $u, v \in U$ be two vertices of interest. 
We know the underlying path $P_{u,v}$ of a fastest temporal path from $u$ to $v$. 
Suppose that $P_{u,v}$ consist of $\ell$ segments $S_{u_i,v_i}$.
Using \cref{claim:FPT-uv-LabelAlmostalledges} we determine the labels of all but $\ell -1$ edges on $P_{u,v}$, with respect to the value of the first edge.
This equals to creating a constraint by setting the duration of the temporal path along $P_{u,v}$ to equal the value $D_{u,v}$.
There are $f(k)$ different paths $P^i_{u,v}$ from $u$ to $v$ in $G$.
We now have to iterate through all of them and add the constraint that the duration of $(P^i_{u,v}, \lambda)$ is greater or equal to $D_{u,v}$.
We repeat the above procedure for all pairs of vertices in $U$.
This results in producing $f(k)$ variables and constraints for our partial ILP.

@@

    It follows that $P$ is a segment.
\end{proof}

@@

From the above it follows that there are at most $f(k)$ (namely $O(|U|^2$) undetermined segments in $G'$ after algorithm finishes going through fastest temporal paths among vertices of interest.

We now continue with guessing fastest temporal paths among all vertices in $U^*$, 
which by similar arguing as for vertices in $U$, gives us $O(k^{k^6})$ guesses.
Besides that we also guess fastest temporal paths of the following form.
Let $S_{u,v} = (u=v_1,v_2, \dots, v_p = v)$ and $S_{w,z} = (w=z_1,z_2, \dots, z_r = z)$ be two segments,
first one between vertices $u,v \in U$, second one between vertices $w, z \in U$, 
where the first one includes $v_2, v_{p-1} \in U^*$, 
and the second one includes $z_2, z_{p-1} \in U^*$.
We want to guess the underlying paths of the fastest temporal paths from $v_2$ to $z_2$,
where the first path goes 
from $v_2$, through $u$, through $w$ and then to $z_2$,
and the second one goes
from $v_2$, through $u$, through $z$ and then to $z_2$.
And two similar paths from $v_2$ to $z_{p-1}$.
We repeat this procedure for all pairs of segments and all vertices from $U^*$ in those segments.
There are $f(k)$ possible paths of this form, since there are $f(k)$ segments in $G'$.
All together we made $f(k)$ new guesses for the structure of the fastest temporal paths.
Each of the previous partial ILP instances is therefore extended into $f(k)$ new ILP instances.

We now do the following.
Let $S^i_{u,v} = (u=v_1,v_2, \dots, v_p = v)$ be a line segment in $G'$, let us
fix a vertex $v_i \in S^i_{u,v} \setminus \{u,v\}$.
Let 
$S^j_{w,z} = (w=z_1,z_2, \dots, z_r = z)$ be another segments in $G'$.
Now we check the matrix $D$ and determine which vertex $z_j \in S^j_{w,z} \setminus \{w,z\}$ 
has the biggest value $D_{v_i,z_j}$.
Let us call this vertex a \emph{split vertex of $v_i$ in the segment $S_{wz}$}.
Denote with $z_{j-1}$ and $z_{j-1}$

Let us first observe some properties.

@@

\begin{claim}\label{claim:FPT-unlabeldPaths-01}

>>>>

Note that we can easily solve the \deltaExact\ by guessing the structure of temporal paths among all pairs of vertices.
But, this would create $f(n,m)$ guesses, and the running time of the algorithm would be too big.
Therefore, we approach the problem using some special properties, which allow us to create $f(k)$ different guesses.
We solve each guess by creating an ILP instance with $f(k)$ variables and $O(n^c)$ constraints, for some $c \in \mathbb{N}$.
The ILP instance is a \textsc{Yes} instance if and only if $D$ is a \textsc{Yes} instance of \deltaExact.
Moreover, from a positive solution of the ILP instance we can construct the desired labeling $\lambda$.
Using results from \cite{Lenstra1983Integer} we know that an ILP instance can be solved in FPT time with respect to the number of variables.
Therefore we can solve each of our ILP instances in $f(k)$ time.

@@

We use the above results in the following way.
Let $u, v \in U$ be two vertices of interest. 
We know the underlying path $P_{u,v}$ of a fastest temporal path from $u$ to $v$. 
Suppose that $P_{u,v}$ consist of $\ell$ segments $S_{u_i,v_i}$.
Using \cref{claim:FPT-uv-LabelAlmostalledges} we determine the labels of all but $\ell -1$ edges on $P_{u,v}$, with respect to the value of the first edge.
This results in creating one constraint, by setting the duration of the temporal path along $P_{u,v}$ to be equal to the value $D_{u,v}$.
There are $O(k^k)$ different paths $P^i_{u,v}$ from $u$ to $v$ in $G$.
We now have to iterate through all of them and add the constraint that the duration of $(P^i_{u,v}, \lambda)$ is greater or equal to $D_{u,v}$.
We repeat the above procedure for all pairs of vertices in $U$.
This results in producing $O(k^{k^3})$ variables and constraints to our partial ILP.

@@

    From all of the above, it follows that $P$ is a segment in $G'$.
\end{proof}

@@

From the above it follows that there are at most $O(|U|^2$) undetermined segments in $G'$ after algorithm finishes going through fastest temporal paths among vertices of interest.

We now continue with guessing fastest temporal paths among all vertices in $U^*$, 
which by similar arguing as for vertices in $U$, gives us $O(k^{k^6})$ guesses.
Besides that we also guess fastest temporal paths of the following form.
Let $S_{u,v} = (u=v_1,v_2, \dots, v_p = v)$ and $S_{w,z} = (w=z_1,z_2, \dots, z_r = z)$ be two segments,
first one between vertices $u,v \in U$, second one between vertices $w, z \in U$, 
where the first one includes $v_2, v_{p-1} \in U^*$, 
and the second one includes $z_2, z_{p-1} \in U^*$.
We want to guess the underlying paths of the fastest temporal paths from $v_2$ to $z_2$,
where the first path goes 
from $v_2$, through $u$, through $w$ and then to $z_2$,
and the second one goes
from $v_2$, through $u$, through $z$ and then to $z_2$.
And two similar paths from $v_2$ to $z_{p-1}$.
We repeat this procedure for all pairs of segments and all vertices from $U^*$ in those segments.
Since there are $O(k^3)$ segments in $G'$,
there are $O(k^6)$ possible paths of this form.
Lastly,
for a segment $S_{u,v} = (u=v_1,v_2, \dots, v_p = v)$
and a vertex of interest $w \in U$,
we guess the underlying paths of the fastest temporal paths from $w$ to $v_2$, through $u$ and through $v$.
Besides that we guess the underlying paths of the fastest temporal paths from $v_2$ to $w$ through $u$ and through $v$.
For a fixed segment $S_{u,v}$ and a fixed vertex of interest $w$ we have $O(4 k^k)$ different possible such paths, therefore we make $O(k^6)$ guesses for these paths.
All together we made $O(k^6)$ new guesses for the structure of all of the above mentioned fastest temporal paths.
Each of the previous partial ILP instances is therefore extended into $O(k^6)$ new ILP instances.

We now do the following.
Let $S^i_{u,v} = (u=v_1,v_2, \dots, v_p = v)$ be a line segment in $G'$, let us
fix a vertex $v_i \in S^i_{u,v} \setminus \{u,v\}$.
Let 
$S^j_{w,z} = (w=z_1,z_2, \dots, z_r = z)$ be another segments in $G'$.
Note that all inner vertices of $S^j_{w,z}$ (\ie vertices in $S^j_{w,z} \setminus \{w,z\}$)
can be reached from $v_i$ by some path that has to go through $w$ oz $z$.
In fact, this is true for all paths starting in any vertex not in $S^j_{w,z}$.
Therefore, the values of the duration of the fastest path from $v_i$ to vertices on 
the segment
$S^j_{w,z}$ have to increase up to the certain point, when traversing the segment coming through $w$ and coming through $z$.
Namely,
let $z_j \in S^j_{w,z} \setminus \{w,z\}$ be the vertex with the biggest value $D_{v_i,z_j}$,
we call this vertex a \emph{split vertex of $v_i$ in the segment $S_{wz}$}.
Then $D_{v_i,z_2} < D_{v_i,z_3} < \dots < D_{v_i,z_j}$ and 
$D_{v_i,z_{p-1}} < D_{v_i,z_{p-2}} < \dots < D_{v_i,z_j}$.
From this it follows that the fastest temporal paths from $v_i$ to $z_2, z_3, \dots, z_{j-1}$ go through $w$,
and 
the fastest temporal paths from $v_i$ to $z_{p-1}, z_{p-2}, \dots, z_{j+1}$ go through $z$.
We now have to guess which vertex $w$ or $z$ is on a fastest temporal path from $v_i$ to $z_j$.
Similarly,
all fastest temporal paths starting at $v_i$ have to go either through $u$ or through $v$,
which also gives us two extra guesses for the underlying path of the fastest temporal path from $v_i$ to $z_j$.
Therefore, all together we have $4$ possibilities on how the underlying path of the fastest temporal path from $v_i$ to $z_j$ 
starts and ends.
For a fixed $S^i_{u,v}, v_i$ and $S^j_{w,z}$ we can find the vertex $z_j$ in polynomial time, 
or determine that this kind of vertex does not exist.
We then make four guesses where we determine how the fastest temporal path from $v_i$ to $z_j$ passes through vertices $u,v$ and $w,z$.
We repeat this procedure for all pairs of segments,
which results in producing $O(k^6)$ new guesses.

Before we can use the above information, let us first prove the following properties.
\begin{claim}\label{claim:FPT-unlabeldPaths-01}
    Suppose that
    $S^i_{u,v}, S^j_{w,z}$ are two segments with $ v_i \in S^i_{u,v}$ and $z_j \in S^j_{w,z}$,
    where $z_j$ is a split vertex of $v_i$ in the segment $S^j_{w,z}$.
    Suppose that the fastest temporal path from $v_i$ to $z_j$ travels through vertices $u$ and $w$.
    Then the fastest temporal path from $v_i$ to any other vertex of $S^j_{w,z}$, that is closer to $w$,
    travels through the same two vertices $u$ and $w$.
    Similarly it holds for the cases when the fastest temporal path travels through $w,v$ or $z,u$ or $z,v$.
\end{claim}
\begin{proof}
    Let $z_\ell$ be a vertex of $S^j_{w,z}$, that is closer to $w$ than $z$ in the segment.
    Let us denote with $P_{v_i,z_j}$ the underlying path of the fastest temporal path from $v_i$ to $z_j$.
    Denote with $P_{v_i,z_j}^\ell$
    the subpath of the fastest temporal path from $v_i$ to $z_j$, that terminates in $z_\ell$.
    We want to show that $P_{v_i,z_j}^\ell$ is an underlying path of a fastest temporal path from $v_i$ to $z_j$.
    Let us observe the following possibilities.
    \\
    First, suppose for the contradiction, that the fastest temporal path from $v_i$ to $z_\ell$ travels through vertices $u$ and $z$.
    Denote this path as $P^1_{v_i,z_\ell}$.
    Then it follows that $d(P^1_{v_i,z_\ell}, \lambda) \leq d(P_{v_i,z_j}^\ell,\lambda)$, which would imply that
    the duration of the temporal path from $v_i$ to $z_j$ using the subpath of $P^1_{v_i,z_\ell}$, would be strictly smaller than the duration of $(P_{v_i,z_j},\lambda)$, which cannot be possible.
    \\
    Second, suppose that the fastest temporal path from $v_i$ to $z_\ell$ travels through vertices $v$ and $w$.
    Denote this path as $P^2_{v_i,z_\ell}$.
    Note that $P_{v_i,z_j}^\ell$ and $P^2_{v_i,z_\ell}$ intersect on a segment $S_{w,z}^j$ from the vertex $w$ to $z_ell$.
    Therefore since
    $d(P^2_{v_i,z_\ell}, \lambda) \leq d(P_{v_i,z_j}^\ell,\lambda)$,
    and since there is unique way to
    extend the path $P^2_{v_i,z_\ell}$ from $z_\ell$ to $z_j$, denote the extended path as $P^j_{v_i,z_\ell}$,
    we get that $d(P^j_{v_i,z_\ell}, \lambda) \leq $$(P_{v_i,z_j},\lambda)$.
    Which implies that $d(P^j_{v_i,z_\ell}, \lambda) = d(P_{v_i,z_j}, \lambda)$.
    Now using the similar argument it follows that $d(P_{v_i,z_j}^\ell, \lambda) = d(P^2_{v_i,z_\ell}, \lambda)$,
    therefore $P_{v_i,z_j}^\ell$ is also a fastest temporal path from $v_i$ to $z_j$.
    \\
    Third, suppose that the fastest temporal path from $v_i$ to $z_\ell$ travels through vertices $v$ and $z$.
    Denote this path as $P^3_{v_i,z_\ell}$.
    Then the duration of the temporal path from $v_i$ to $z_j$ using the subpath of $P^3_{v_i,z_\ell}$, would be strictly smaller than the duration of $(P_{v_i,z_j},\lambda)$, which cannot be possible.
\end{proof}

We proceed with adding constraints to our ILP instance, by iterating through all pairs of vertices.
We already studied the case of fastest paths starting and ending in a vertex of $U$,
so now we have to study all other cases.

Let us start with the case of determining the fastest path from $u$ to $x$, where $u \in U$ and $x \in V(G') \setminus U$.
We have two possibilities:
\begin{enumerate}[(i)]
    \item Vertex $x$ is on a fastest temporal path from $u$ to some other vertex of interest $v$.
    Suppose that $x$ lies on a segment $S_{w,z}$.
    Then by \cref{claim:FPT-uv-LabelAlmostalledges} we know that the fastest temporal path from $u$ to $x$ is a subpath of the fastest temporal path from $u$ to $v$,
    which hast to be of the duration $D_{u,x}$. This results in adding an equality constraint to our ILP.
    We now have to iterate through all others temporal paths from $u$ to $x$ and add the constraint to ILP that states that their duration is not smaller than the value $D_{u,x}$.
    To reach $x$ from $u$, we have to first travel from $u$ to $w$ (resp.~$z$), and then travel through $S_{u,w}$ to $x$.
    There are $O(k^k)$ different paths to reach $w$ (resp.~$z$), which are then uniquely extended to $x$.
    Therefore we add $O(k^k)$ inequality constraints for all these paths.

    \item Vertex $x$ is not on a fastest temporal paths from $u$ to any $v \in U$.
    Suppose that $x$ lies on a segment $S_{w,z} = (w,z_1,2_3, \dots, z_p, z)$.
    Since we know the underlying path of the fastest temporal path from $u$ to $z_1$ through $w$ and from $u$ to $z_p$ through $z$,
    we have to only determine how we reach vertex $x$ from the segment (\ie will the fastest path go through $w$ or through $z$).
    This is done,
    by checking the distances from $u$ to $z_i$, for $z_i \in S_{w,z} \setminus \{w,z\}$,
    and determining the split vertex $w^*$ from $u$ on the segment $S_{w,z}$.
    If $x \neq w^*$, we know which part of the segment is used to reach $x$.
    We set this path as the fastest temporal path, that must satisfy the equalities and for all other temporal paths
    (\ie paths from $u$ to $z$ and then to $x$, and paths from $u$ to $w$ and then to $x$),
    we add the inequalities constraints.
    If $x = w^*$, we have two possibilities, we check the values of each one and set the minimum of them, to equal the $D_{u,x}$ (if possible, if one is to small we immediately reject),
    and again set all other possibilities to the inequality constraint.
    There are $O(k^3)$ different paths from $u$ to $w$ (resp.~$z$), therefore we add $O(k^k)$ inequality constraints for all these paths.   
\end{enumerate}

Now we proceed with the case of determining the fastest path from $x$ to $u$, where $x \in V(G') \setminus U$ and $u \in U$.

Next is the case of determining the fastest path from $x$ to $y$, where $x,y \in V(G') \setminus U$.

\section{FPT, Old proof -- need to move useful stuff above}
Let us first observe some properties.

@@

\begin{claim}\label{claim:FPT-unlabeldPaths-012}
